Lossy compression of DNA sequencing quality data

### 2015-05-03 manuscript/q-compression.tex
Continued revisions ...

We evaluate effects of information loss from quality value compression
on quality control steps of trimming and read filtering, which were
performed using Sickle~\citep{sickle}, and make comparison to
uncompressed data. Sickle starts at the ends of the sequence and uses
a sliding window ($0.1 \times$ the read length) to find locations
where the window's average quality falls below a given threshold (20
by default). The sequence is then trimmed from this position to the
respective end. If the trimmed sequence length is less than a certain
threshold (20 by default), then the sequence is discarded.

@@

We compare the \textsc{mse} versus bits/base-call of the
\textit{Rhodobacter sphaeroides}, \textit{Homo sapiens},
\textit{Escherichia coli}, and \textit{Mus musculus} data sets (Figure
\ref{fig:mse_vs_bpbp}). We only include the fragment libraries for the
\textit{Rhodobacter sphaeroides}, and \textit{Homo sapiens} data sets,
but the additional short-jump library results are available in the
Supplemental Table 1. Storing the uncompressed quality values requires
1 byte per base-call because they are stored in \textsc{ascii} format
and is denoted by the dotted black asterisk in the figure. The
lossless compression of each data set using \textsc{bz}ip2 ranges from
2.19--3.10 bits/base-call and is denoted by the colored asterisks on
the abscissa. The compression methods tend to cluster together across
the different data sets. Across all data sets, the 0-degree polynomial
regression, profile encodings, and \textsc{q}ual\textsc{c}omp have the
lowest bits/base-call.

\textsc{q}ual\textsc{c}omp with the rate parameter set to 100
bits/read has the lowest \textsc{mse}, but requires
$\sim$10--15$\times$ more storage than the profile encoding methods
for only a $\sim$2--3$\times$ reduction in \textsc{mse}. When
\textsc{q}ual\textsc{c}omp's rate parameter is set to match the
profile encoding methods, \textsc{q}ual\textsc{c}omp performs slightly
worse in terms of \textsc{mse}. In the \textit{Rhodobacter
  sphaeroides} fragment library, \textsc{q}ual\textsc{c}omp with rate
10 bits/read (0.099 bits/bp) has a \textsc{mse} of 17.29. Using
256-profile encoding requires less storage (0.079 bits/bp) and has a
lower \textsc{mse} (11.85).

>>>>

We evaluate effects of information loss from quality value compression
on quality control steps of trimming and read filtering, which were
performed using Sickle~\citep{sickle}, and make comparison to
uncompressed data. Sickle starts at the ends of the read and uses
a sliding window ($0.1 \times$ the read length) to find locations
where the mean quality in the window falls below a given threshold (20
by default). The sequence is then trimmed from this position to the
respective end. If the trimmed sequence length is less than a certain
threshold (20 by default), then the sequence is discarded.

@@

We compare the \textsc{mse} versus bits/base-call of the
\textit{Rhodobacter sphaeroides}, \textit{Homo sapiens},
\textit{Escherichia coli}, and \textit{Mus musculus} data sets (Figure
\ref{fig:mse_vs_bpbp}). Here we present the fragment libraries for the
\textit{Rhodobacter sphaeroides}, and \textit{Homo sapiens} data sets,
but the short-jump library results are available in the Supplemental
Table 1. Storing the uncompressed quality values requires 1
byte/base-call because they are stored in \textsc{ascii} format and is
denoted by the dotted black asterisk in the figure. The lossless
compression of each data set using \textsc{bz}ip2 ranges from
2.19--3.10 bits/base-call and is denoted by the colored asterisks on
the abscissa. The compression methods tend to cluster together across
the different data sets. Across all data sets, the 0-degree polynomial
regression, profile encodings, and \textsc{q}ual\textsc{c}omp have the
lowest bits/base-call.

\textsc{q}ual\textsc{c}omp with the rate parameter set to 100
bits/read has the lowest \textsc{mse}, but requires
$\sim$10--15$\times$ more storage than the profile encoding methods
for only a $\sim$2--3$\times$ reduction in \textsc{mse}. When rate
parameter of \textsc{q}ual\textsc{c}omp is set to match the profile
encoding methods, \textsc{q}ual\textsc{c}omp performs slightly worse
in terms of \textsc{mse}. For the \textit{Rhodobacter sphaeroides}
fragment library, \textsc{q}ual\textsc{c}omp with rate 10 bits/read
(0.099 bits/bp) has a \textsc{mse} of 17.29, whereas 256-profile
encoding only requires 0.079 bits/bp and has a \textsc{mse} of 11.85.
